Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-04-01 manuscript.tex
add numbers

enhancers (x.x \%) are predicted to be specific to pre-migratory neural crest by both models.\par

@@

\begin{figure}[b]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
 \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
    for $DeepNeuralTube_{organoid}$ (top two panels) and $DeepNeuralTube_{embryo}$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
    $DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
    enhancer (chr ...; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

@@

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and SUPPLEMENTAL FIG). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and SUPPLEMENTAL FIG). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.} \\
    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells). (Continues next page).
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
  }
\end{figure}

@@

From these results it seems that a relatively simplistic code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains. Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig. 4i}). Surprisingly, this very simple classifier reached high performance with area under the ROC of X and area under the precision recall curve of X (\textbf{Fig. 4i}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models. Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G \textgreater A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par

@@

Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube_{organoid}$ and
$DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each topic and generated a code table based on the
identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
of these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 6b}).\par

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
tube organoids and the head of a four weeks post conception human embryo with almost one-to-one correspondence based on TFBS importances and TF expression. Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average area under the ROC of X and area under the precision-recall curve of X. Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). Finally, we used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. We used at least three TFBS in the design process for each state. In case less than three distinct TFBS had a positive coefficient for a state, we implanted multiple copies of each TFBS to come to a total of at least four TFBS. The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par

>>>>

enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par

@@

\begin{figure}[b]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
 \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
    for $DeepNeuralTube_{organoid}$ (top two panels) and $DeepNeuralTube_{embryo}$ (bottom two panels) and enhancer-reporter
    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
    $DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
    constitutive enhancer that serves as electroporation control.
  }
\end{figure}

@@

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and SUPPLEMENTAL FIG). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and SUPPLEMENTAL FIG). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

@@

\afterpage{\clearpage}
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
  \caption{
    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.} \\
    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
    and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells). (Continues next page).
  }
\end{figure}
\afterpage{\clearpage}

\begin{figure}[b!]
  \ContinuedFloat
  \caption*{\rule{\linewidth}{0.4pt}}
  \caption{
    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
  }
\end{figure}

@@

From these results it seems that a relatively simplistic code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains. Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig. 4i}). Surprisingly, this very simple classifier reached high performance with area under the ROC of 0.82 and area under the precision recall curve of 0.80 (\textbf{Fig. 4i}).\par

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models. Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par

@@

Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube_{organoid}$ and
$DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each topic and generated a code table based on the
identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
of these patterns is highly correlated across both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
tube organoids and the head of a four weeks post conception human embryo with almost one-to-one correspondence based on TFBS importances and TF expression. Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average area under the ROC of 0.77 and area under the precision-recall curve of 0.39. Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). Finally, we used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. We used at least three TFBS in the design process for each state. In case less than three distinct TFBS had a positive coefficient for a state, we implanted multiple copies of each TFBS to come to a total of at least four TFBS. The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par
